\textbf{Linear Algebra -- Part II}\\A summary for MIT 18.06SC

### 2020-07-24 skills/math/linear_algebra/linear_algebra_summary_p2.tex
revision

For notational convenience, we use the shorthands $\mA$ for the column space $C(\mA)$ and $\mA^{\perp}$ for the nullspace $N(\mA)$ in the rest of this section. Though $\mA$ usually denotes a matrix, the meaning of $\mA$ should be clear given the context. For example, $\rank(\mA)$ refers the rank of matrix $\mA$, while $\dim(\mA)$ refers to the dimension of the column space $C(\mA)$. Specifically, for a vector $v$, $v \in \mA$ means that the vector $v$ belongs to the column space $C(\mA)$.

@@

	On the other hand, if the vector $x \in (\mA^T \mA)^{\perp}$, we have
	\[ \mA^T \mA x = 0.\]
	Then, the vector $\mA x \in  \mA^{T,\perp}$. By the definition of matrix-to-vector multiplication, we also have $\mA x \in \mA$.  By the Theorem~\ref{thm:ortho}, the subspaces $\mA^{T,\perp} \perp \mA$. Therefore, $\mA x = 0$, and we obtain that $x \in \mA^{\perp}$.
	\vspace{0.2cm}

@@

\subsection{Projection matrices and least squares}
Let $y \in \bbR^n$ be a vector and $\mX \in \bbR^{n \times (k+1)}$ be a design matrix. We propose the linear regression model as
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)^T$ are regression coefficients and $\epsilon$ is the noise. The least-squares estimate of $\beta$ is the minimizer of the squared loss; i.e.,
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the Euclidean norm.  The projection of $y$ onto $\mX$ minimizes the Euclidean distance from $y$ to the column space $\mX$. Hence, the vector $\mX \hat \beta_{LS}$ is the projection of $y$ onto the column space $\mX$. Therefore, we may use projection to solve the least squares problem.

@@

	By properties~\ref{propdiag} and \ref{prop2}, the determinant of a nonsingular $\mA'_i$ follows the formula \[(-1)^{N(\alpha_1,...,\alpha_n)}  a_{1\alpha_1} ... a_{n \alpha_n},\]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$, $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of swaps from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. The determinant of $\mA$ follows by the summation of the nonsingular matrices' determinants.
\end{proof}

@@

\begin{proof}
	First, we re-write the characteristic polynomial $P(\lambda)$ of $\mA$ using the factorization form,
	\begin{equation}\label{eq:poly}
		P(\lambda) = (-1)^n (\lambda - \lambda_1)\cdots (\lambda - \lambda_n), 
	\end{equation}   
	where $\lambda_i$ are the eigenvalues of $\mA$, for all $ i \in [n]$. In equation~\eqref{eq:poly}, the coefficient for $\lambda^{n-1}$ is equal to $(-1)^n\sum_{i=1}^n \lambda_i$. Comparing the equation~\eqref{eq:poly} with Definition~\ref{def:charac}, we obtain $\tr(\mA) = \sum_{i=1}^n \lambda_i$. Similarly, the constant term in equation~\eqref{eq:poly} is $(-1)^n \prod_{i=1}^n \lambda_i $, while the constant term in Defintion~\ref{def:charac} is $(-1)^n \det(\mA)$.  Therefore, we obtain $\prod_{i=1}^n \lambda_i = \det(\mA)$.
\end{proof}

@@

\begin{proof}
	Since $\mA$ is invertible, the eigenvalue $\lambda_i \neq 0$, for all $  i \in [n]$. Multiplying $x_i$ on both sides of the equation $\mA^{-1} \mA = \mI_n$ yields
	\[ \mA^{-1} \mA x_i = x_i \quad \Rightarrow\quad  \mA^{-1} x_i = {\lambda_i}^{-1} x_i, \quad \text{for all } i \in [n]. \]
Therefore, $(\lambda_i^{-1},x_i)$ is the eigenvalue-eigenvector pair of $\mA^{-1}$, for all $i \in [n]$.
\end{proof}

@@

\begin{proof}
	Suppose the square matrix $\mA \in \bbR^{n \times n}$ has $m$ distinct eigenvalues $\lambda_1,...,\lambda_m$ with associated eigenvectors  $v_1,...,v_m$. Consider a linear combination $c_1 v_1 + ... + c_m v_m = 0$, where $c_i \in \bbR$, for all $i \in [m]$.	
	Then, we have
	\begin{equation}\label{eq:indep1}
		\mA(c_1 v_1 + ... + c_m v_m ) = c_1 \lambda_1 v_1 + ... + c_m \lambda_m v_m = 0,
		\end{equation}
		and
		\begin{equation}\label{eq:indep2}
		\lambda_1 (c_1 v_1 + ... + c_m v_m ) = c_1 \lambda_1 v_1 + ... + c_m \lambda_1 v_m = 0. 
	\end{equation}
	Subtracting the equations~\eqref{eq:indep1} from \eqref{eq:indep2} yields
	\begin{equation}\label{eq:indep3}
		 c_2 (\lambda_1 - \lambda_2) v_2 + \cdots + c_m (\lambda_1 - \lambda_m) v_m  = 0.
	\end{equation}
	Consider the new linear combination $ c_2 (\lambda_1 - \lambda_2) v_2 + \cdots + c_m (\lambda_1 - \lambda_m) v_m = 0$, and repeat the steps from equation~\eqref{eq:indep1} to equation~\eqref{eq:indep3}. After $m$ iterations, we have 
	\[ c_m(\lambda_1 - \lambda_m)(\lambda_2 - \lambda_m)\cdots (\lambda_{m-1} - \lambda_m) v_m = 0.    \]
	 Since $\lambda_1 ,..., \lambda_m$ are distinct, and $v_m$ is a nonzero vector, the coefficient $c_m = 0$.  Plugging $c_m = 0$ back to previous steps, we have $c_1 = \cdots = c_m = 0$. Therefore, $v_1,...,v_m$ are independent.
\end{proof}

@@

\begin{defn}[Antisymmetric matrices]
	The matrix $\mA$ is an antisymmetric matrix if 
	\[\mA^T = - \mA. \] 
\end{defn}

@@

	Let  $\mA$  be a symmetric  matrix with real entries only. Suppose $\lambda$  is a complex eigenvalue of $\mA$ with  a nonzero imaginary part, and $x$ is the eigenvector associated with $\lambda$. Since all the entries of $\mA$ are real, the eigenvector $x$ is also a complex vector with a nonzero imaginary part. Let $\overline \lambda$ and $\overline x$ denote the conjugate eigenvalue and eigenvector, respectively. Then,  we have
	\begin{equation}\label{eq:symm}
		\overline x^T \mA x = \overline x^T \lambda x,
	\end{equation} 
	and
	\begin{equation}\label{eq:symm2}
		x^T \mA \overline x = x^T \overline \lambda \overline x.
	\end{equation}
	 By the symmetry of $\mA$, we have
	\[ \overline x^T \mA x  = x^T \mA^T \overline x = x^T \mA \overline x .\] 
	 Subtracting the equation~\eqref{eq:symm2} from equation~\eqref{eq:symm} yields	\[  0 = ( \overline x^T \mA x -  x^T \mA \overline x ) = (\lambda - \overline \lambda) x^T \overline x.  \]
	Since that $x^T \overline x$ is a nonzero real number, the imaginary part of $\lambda$ is equal to 0, which contradicts the assumption that $\lambda$ has a nonzero imaginary part. Therefore, the eigenvalues of $\mA$ are real.

@@

\begin{proof}
	By Theorems~\ref{thm:symmortho} and \ref{thm:symmrepeat}, for any symmetric matrix $\mA \in \bbR^{n \times n}$, $\mA$ have $n$ orthogonal eigenvectors with possibly repeated eigenvalues. Therefore, the matrix $\mA$   is able to be diagonalized as equation~\eqref{eq:defdiag}.
\end{proof}

@@

\begin{proof}
Since $\mA$ has $n$ independent eigenvectors, the matrix $\mA$ is diagonalizable.
	Let $\mA =\mS^{-1} \Lambda \mS$, where $\Lambda, \mS$ are defined the same as Definition~\ref{def:diagonalization}. For all integer $k\geq 0$, we have $\mA^k = \mS^{-1} \Lambda^{k} \mS$. Then,
	\[ \lim_{k \rightarrow +\infty} \mA^k = \lim_{k \rightarrow +\infty} \mS^{-1} \Lambda^{k} \mS  =  \lim_{k \rightarrow +\infty} \mS^{-1} \begin{bmatrix}
		|\lambda_i|^k & \cdots &0\\
		\vdots & &\vdots\\
		0 & \cdots & |\lambda_n|^k
	\end{bmatrix} \mS  = 0 \quad \Leftrightarrow \quad  |\lambda_i| < 1, \text{ for all } i \in [n]. \]
\end{proof}

>>>>

For notational convenience, we use the shorthands $\mA$ for the column space $C(\mA)$ and $\mA^{\perp}$ for the nullspace $N(\mA)$ in the rest of this section. Though $\mA$ usually denotes a matrix, the meaning of $\mA$ should be clear given the context. For example, $\rank(\mA)$ refers to the rank of matrix $\mA$, while $\dim(\mA)$ refers to the dimension of the column space $C(\mA)$. Specifically, for a vector $v$, $v \in \mA$ means that the vector $v$ belongs to the column space $C(\mA)$.

@@

	On the other hand, if the vector $x \in (\mA^T \mA)^{\perp}$, we have
	\[ \mA^T \mA x = 0.\]
	Then, the vector $\mA x \in  (\mA^{T})^{\perp}$. By the definition of matrix-to-vector multiplication, we also have $\mA x \in \mA$.  By Theorem~\ref{thm:ortho}, the subspaces $(\mA^{T})^{\perp} \perp \mA$. Therefore, $\mA x = 0$, and we obtain that $x \in \mA^{\perp}$.
	\vspace{0.2cm}

@@

\subsection{Projection matrices and least squares}
Let $y \in \bbR^n$ be a vector and $\mX \in \bbR^{n \times (k+1)}$ be a design matrix. We propose the linear regression model as
\[ y = \mX \beta +\epsilon,\]
where $\beta = (\beta_0,\beta_1,...,\beta_k)^T$ are regression coefficients and $\epsilon$ is the noise. The least-squares estimate of $\beta$ is the minimizer of the squared loss; i.e.,
\[  \hat \beta_{LS} = \argmin_{\beta \in \bbR^{k+1}} \onorm{y - \mX \beta}^2, \]
where $\onorm{\cdot}$ is the Euclidean norm.  The projection of $y$ onto $\mX$ minimizes the Euclidean distance from $y$ to the column space $\mX$. Hence, the vector $\mX \hat \beta_{LS}$ is the projection of $y$ onto the column space $\mX$. Therefore, we may use projection to solve the least-squares problem.

@@

	By properties~\ref{propdiag} and \ref{prop2}, the determinant of a nonsingular $\mA'_i$ follows the formula \[(-1)^{N(\alpha_1,...,\alpha_n)}  a_{1\alpha_1} ... a_{n \alpha_n},\]
	where $(\alpha_1, \alpha_2, ...,\alpha_n)$ is the permutation of $(1,2,...,n)$, $\tP$ is the collection of all possible  $(\alpha_1,...,\alpha_n)$, and $N(\alpha_1,...,\alpha_n)$ is the number of swaps from $(\alpha_1,\alpha_2,...,\alpha_n)$  to $(1,2,...,n)$. The determinant of $\mA$ is the sum of the nonsingular matrices' determinants.
\end{proof}

@@

\begin{proof}
	First, we re-write the characteristic polynomial $P(\lambda)$ of $\mA$ using the factorization form,
	\begin{equation}\label{eq:poly}
		P(\lambda) = (-1)^n (\lambda - \lambda_1)\cdots (\lambda - \lambda_n), 
	\end{equation}   
	where $\lambda_i$ are the eigenvalues of $\mA$, for all $ i \in [n]$. In equation~\eqref{eq:poly}, the coefficient for $\lambda^{n-1}$ is equal to $(-1)^n\sum_{i=1}^n \lambda_i$. Comparing equation~\eqref{eq:poly} with Definition~\ref{def:charac}, we obtain $\tr(\mA) = \sum_{i=1}^n \lambda_i$. Similarly, the constant term in equation~\eqref{eq:poly} is $(-1)^n \prod_{i=1}^n \lambda_i $, while the constant term in Defintion~\ref{def:charac} is $(-1)^n \det(\mA)$.  Therefore, we obtain $\prod_{i=1}^n \lambda_i = \det(\mA)$.
\end{proof}

@@

\begin{proof}
	Since $\mA$ is invertible, the eigenvalue $\lambda_i \neq 0$, for all $  i \in [n]$. Multiplying $x_i$ on both sides of the equation $\mA^{-1} \mA = \mI_n$ yields
	\[ \mA^{-1} \mA x_i = x_i \quad \Rightarrow\quad  \mA^{-1} x_i = {\lambda_i}^{-1} x_i, \quad \text{for all } i \in [n]. \]
Therefore, $(\lambda_i^{-1},x_i)$ is the $i$-th eigenvalue-eigenvector pair of $\mA^{-1}$, for all $i \in [n]$.
\end{proof}

@@

\begin{proof}
	Suppose the square matrix $\mA \in \bbR^{n \times n}$ has $m$ distinct eigenvalues $\lambda_1,...,\lambda_m$ with associated eigenvectors  $v_1,...,v_m$. Consider a linear combination $c_1 v_1 + ... + c_m v_m = 0$, where $c_i \in \bbR$, for all $i \in [m]$.	
	Then, we have
	\begin{equation}\label{eq:indep1}
		\mA(c_1 v_1 + ... + c_m v_m ) = c_1 \lambda_1 v_1 + ... + c_m \lambda_m v_m = 0,
		\end{equation}
		and
		\begin{equation}\label{eq:indep2}
		\lambda_1 (c_1 v_1 + ... + c_m v_m ) = c_1 \lambda_1 v_1 + ... + c_m \lambda_1 v_m = 0. 
	\end{equation}
	Subtracting equations~\eqref{eq:indep1} from \eqref{eq:indep2} yields
	\begin{equation}\label{eq:indep3}
		 c_2 (\lambda_1 - \lambda_2) v_2 + \cdots + c_m (\lambda_1 - \lambda_m) v_m  = 0.
	\end{equation}
	Consider the new linear combination $ c_2 (\lambda_1 - \lambda_2) v_2 + \cdots + c_m (\lambda_1 - \lambda_m) v_m = 0$, and repeat the steps from equation~\eqref{eq:indep1} to equation~\eqref{eq:indep3}. After $m$ iterations, we have 
	\[ c_m(\lambda_1 - \lambda_m)(\lambda_2 - \lambda_m)\cdots (\lambda_{m-1} - \lambda_m) v_m = 0.    \]
	 Since $\lambda_1 ,..., \lambda_m$ are distinct, and $v_m$ is a nonzero vector, the coefficient $c_m = 0$.  Plugging $c_m = 0$ back to previous steps, we have $c_1 = \cdots = c_m = 0$. Therefore, $v_1,...,v_m$ are independent.
\end{proof}

@@

\begin{defn}[Antisymmetric matrices]
	A matrix $\mA$ is antisymmetric if 
	\[\mA^T = - \mA. \] 
\end{defn}

@@

	Let  $\mA$  be a symmetric  matrix with real entries only. Suppose $\lambda$  is a complex eigenvalue of $\mA$ with  a nonzero imaginary part, and $x$ is the eigenvector associated with $\lambda$. Since all the entries of $\mA$ are real, the eigenvector $x$ is also a complex vector. Let $\overline \lambda$ and $\overline x$ denote the conjugate eigenvalue and eigenvector, respectively. Then,  we have
	\begin{equation}\label{eq:symm}
		\overline x^T \mA x = \overline x^T \lambda x,
	\end{equation} 
	and
	\begin{equation}\label{eq:symm2}
		x^T \mA \overline x = x^T \overline \lambda \overline x.
	\end{equation}
	 By the symmetry of $\mA$, we have
	\[ \overline x^T \mA x  = x^T \mA^T \overline x = x^T \mA \overline x .\] 
	 Subtracting equation~\eqref{eq:symm2} from equation~\eqref{eq:symm} yields	\[  0 = ( \overline x^T \mA x -  x^T \mA \overline x ) = (\lambda - \overline \lambda) x^T \overline x.  \]
	Since that $x^T \overline x$ is a nonzero real number, the imaginary part of $\lambda$ is equal to 0, which contradicts the assumption that $\lambda$ has a nonzero imaginary part. Therefore, the eigenvalues of $\mA$ are real.

@@

\begin{proof}
	By Theorems~\ref{thm:symmortho} and \ref{thm:symmrepeat}, for any symmetric matrix $\mA \in \bbR^{n \times n}$, $\mA$ has $n$ orthogonal eigenvectors with possibly repeated eigenvalues. Therefore, the matrix $\mA$   is able to be diagonalized as equation~\eqref{eq:defdiag}.
\end{proof}

@@

\begin{proof}
Since $\mA$ is diagonalizable, let $\mA =\mS^{-1} \Lambda \mS$, where $\Lambda, \mS$ are defined the same as Definition~\ref{def:diagonalization}. For all integer $k\geq 0$, we have $\mA^k = \mS^{-1} \Lambda^{k} \mS$. Then,
	\[ \lim_{k \rightarrow +\infty} \mA^k = \lim_{k \rightarrow +\infty} \mS^{-1} \Lambda^{k} \mS  =  \lim_{k \rightarrow +\infty} \mS^{-1} \begin{bmatrix}
		|\lambda_i|^k & \cdots &0\\
		\vdots & &\vdots\\
		0 & \cdots & |\lambda_n|^k
	\end{bmatrix} \mS  = 0 \quad \Leftrightarrow \quad  |\lambda_i| < 1, \text{ for all } i \in [n]. \]
\end{proof}
